---
title: 'Tipping Points and Critical Thresholds in Model Systems'
author: "Alistair Seddon"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
header-includes: 
 - \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library(png) 

```

# Overview
- What do we mean by dynamical behaviour?
- What are critical transitions in model systems?
- What are some characteristic behaviours of critical transitions in different datasets?
- What are the implications for these ideas in palaeoecological research?

# Critical Transitions in Nature and Society
- Disclaimer: I am **not** a theoretical ecologist
- I became interested in complex systems as a palaeoecologist, and became interested in how to apply these ideas to palaeoecological datasets
- Much of the following slides are based on Scheffer's (2009) book, taking the examples from the back of the book.

```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img1_path <- "images/Scheffer2009.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
#attr(img1, "info")
knitr::include_graphics(img1_path)
```

# Dynamical Systems (A very brief introduction)
- Fixed rule (e.g. mathematical model) determines time dependence of a point in geometric space
- Dynamical system theory used to explain the interaction of lots of processes
- Examples of dynamical systems include: Fish populations, a pendulum swinging.

# Characteristics of dynamical systems 
- State depends on a balance of processes
- **Stability** and **equilibirum**
- Equilibrium in the Earth's temperature
- Carrying capacity in a population: means the population tends of equilibrium
- Stability landscapes an important concept
- Otherwise known as an *attractor* or *stable state*
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img2_path <- "images/stability_equilibrium.png"
knitr::include_graphics(img2_path)
```

<!-- Notes: Dynamical systems: branch of mathematics describing the interaction of different processes- Earth is a dynamical system as it is the result of different processes.

Population density is a function of birth rates and death rates, if you increase pop densitiy above the carrying capacity, fewer resources and die off, if you decrease, more resources and return to equilibrium carrying capacity
 -->
 
# Alternative equilibria
- In a population we can also have something called an Allee effect
- At lower population densities, it can be harder for individuals to find food, or each other. If you reach this low number the population tends to crash.
- The result is that you now have *stable* and *unstable* regions, where populations are *repelled* or *attracted* to depedning on where you are in the stability plot.
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img3_path <- "images/allee.png"
knitr::include_graphics(img3_path)
```

# Other characteristics of dynamical systems (1)
- Seemingly random patterns can actually be deterministic
- Example 1: provides an example of a completely noisy system vs. a choatic one
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img4_path <- "images/sugihara_may_1.png"
knitr::include_graphics(img4_path)
```

<!-- - A(t+1) = r A * (1-A) -->

<!-- ```{r Chaos vs. Noise, echo = TRUE} -->
<!-- ### Noisy system -->
<!-- t <- 1:100 -->
<!-- x1 <- runif(length(t), 0.2, 0.8 ) -->

<!-- ### Choatic system -->
<!-- r <- 3.3 -->
<!-- starting.x2 <- runif(1,0.1,0.9) -->

<!-- x2 <- rep(NA, length(t)) -->
<!-- x2[1] <- starting.x2 -->

<!-- for(i in 2: length(t)) { -->
<!--   x2[i] <- r* x2[i-1] * (1 -x2[i-1]  )  + runif(1,0,0.2) -->
<!-- } -->
<!-- x2[is.infinite(x2)] <- 0 -->

<!-- eq1Tib <- tibble(Time = t, x1 = x1, x2 = x2 ) -->

<!-- ``` -->

<!-- # Other characteristics of dynamical systems (1) -->
<!-- - Plot these data -->
<!-- ```{r Chaos vs. Noise Plot, echo = TRUE} -->
<!-- fig1 <- eq1Tib %>%  -->
<!--   pivot_longer(c(x1, x2), names_to = "model",  values_to = "Pop.density") %>%  -->
<!--   ggplot(aes(x = Time, y = Pop.density)) + -->
<!--   geom_point(aes(col = model)) + -->
<!--   geom_line(col = "black") + -->
<!--   facet_grid(model ~.) -->

<!-- fig1 -->
<!-- ``` -->


# Characteristics of dynamical systems
- Deterministic chaos
- Seemingly random patterns can actually be deterministic (chaotic)
- Lagged-coordinate embedding can be used to predict future timescales, and is another way to identify attractors

```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img5_path <- "images/sugihara_may_2.png"
knitr::include_graphics(img5_path)
```

<!-- ```{r Lagged Coordinate Embedding, echo = TRUE} -->
<!-- x1.1 <- c(x1[2:length(x1)], NA) -->
<!-- x2.1 <- c( x2[2:length(x2)], NA) -->

<!-- lags.x1 <- tibble(Time = t, xt = x1, xt1 = x1.1, model = "x1") -->
<!-- lags.x2 <- tibble(Time = t, xt = x2, xt1 = x2.1, model = "x2") -->

<!-- lags <- bind_rows(lags.x1, lags.x2) -->

<!-- fig2 <- lags %>% -->
<!--   ggplot(aes(x = xt1, y = xt)) + -->
<!--   geom_point(aes(col = model)) + -->
<!--   facet_grid(model ~.) -->

<!-- fig2 -->

<!-- ``` -->
<!-- One interesting thing about lagged coordinate embedding is that we can use the previous timestep to predict what is going to happen next, without understanding the actual equations or processes that govern the system… taking advantage of these kind of properties of dynamical systems is something that we can come back to later. -->
<!-- Plotting these values (t+1 vs t) is another way of finding attractors or equilibrium. Where the points converge demonstrates the attractors, at these points can quite confidently predict what the next value is going to be. At other points (at the singularities), it’s more difficult (singularities) -->

# Characteristics of dynamical systems
- The butterfly-shaped Lorenz attractor is an example: the path of the marble or airplane is constrained to certain limits and has a charcteristic trajectory, but it never quite follows the exact same path twice.
- *Elk population example*. Carrying capacity of 100. A great many factors are responsible for changing the elk population: though it never goes extinct--nor increases to infinity--the simple swing circle from 120 to 80 and back again is disrupted and the peaks and valleys are a bit different each time.

```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img6_path <- "images/Lorenz_attractor.png"
knitr::include_graphics(img6_path)
```

# Characteristics of dynamical systems
Attractors can also be cycles, e.g. when one system driven by strong periodic forcing
- e.g. plankton cycle driven by seasonal variability
- Dependent on two frequencies interacting

Or cycles driven by internal dynamics, the interaction of fast and slow processes
- e.g. beetle outbreaks in forests.

# Exercise 1
**Aims**:
- Get familiar with a simple population and explore dynamical behaviour in this model
- Refer to the hand-out for the instructions, and the Rscript `resilience_exercise.R` for the code

# Exercise 1: Code
```{r Task1 A simple population model}

# Here is the model describing population density of a species. Population density (A) at time t + 1 is dependent on the population densitiy at time t, multiplied by the birth rate (r). The second part of the equation (1-At) reduces the size of the population at higher values of A.

# A[t+1] = r*A[t] * (1 - A[t])

# a) Investigate the dynamical behaviour of the system for different levels of r. What happens as increase r past a certain point?
# b) Use the lagged coordinate embedding technique as another way to show that the system is deterministic
# a) 
t <- 1:100
r <- 3 # () Max = 4)
starting.A <- runif(1,0.1,0.9) # sample the starting value

A <- rep(NA, length(t))
A[1] <- starting.A

for(i in 2: length(t)) {
  A[i] <- r* A[i-1] * (1 -A[i-1]  )
}
A[is.infinite(A)] <- 0

modelResult<- tibble(Time = t, A = A)
modelResult %>%
ggplot(aes(x = Time, y = A)) +
  geom_point() +
  geom_line(col = "black")

#
# b) Exploring the lagged coordinate embedding
A.t1 <- c(A[2: length(t)], NA)
lagPlot <- tibble(A.t1 = A.t1, A = A)
lagPlot %>%
ggplot(aes(x= A.t1, y = A)) +
  geom_point()

```

# Exercise 2: Ecological equilibria and rates of population growth
**Aims**:
- Explore the effects of changing the carrying capacity and growth rates in the model
- Identify alternative stable states with relation to an Allee effect
- Refer to the hand-out for the instructions, and the Rscript `resilience_exercise.R` for the code

```{r Task2 A simple population model}

# This is a slightly more complicated model with the carrying capacity added, demonstrating logistic growth for the producers (e.g. Plants). Population increase (dA/dt) is a function of population density (A) and two parameters, K= Carrying capacity, and r= growth rate

t <- 1:100
K <- 75 # Carrying capacity, range 50-100
r <- 0.3 # Growth rate, range 0-1

starting.A <- runif(1,0,100) # sample the starting value

A <- rep(NA, length(t))
A[1] <- starting.A

dA.dt <- rep(NA, length(t)) # Rate of population growth for timestep 1

for(i in 1: (length(t)-1)) {
  dA.dt[i] <-  r* A[i] * (1 -(A[i]/K) )
  A[i+1] <- A[i]+ dA.dt[i]
  }

A[is.infinite(A)] <- 0

modelResult<- tibble(Time = t, dA.dt = dA.dt, A = A)

popDensityPlot <-modelResult %>%
ggplot(aes(x = Time, y = A)) +
  geom_point() +
  geom_line(col = "black") +
  ylim(0,100)

popDensityPlot
```

a) Use the code above to experiment how changing the different the parameters (e.g. growth rates) influences the shape of the growth. What happens when the starting value is greater/ lower than the carrying capacity?

b) Make a plot to show how rate of population growth (dA/dt) changes as a function of population density. Specifically, what happens to the rate of population growth as we approach an equilibrium?

```{r Task2b A simple population model- rate of population growth}
 modelResult %>%
   ggplot(aes(x = A, y = dA.dt)) +
  geom_point() +
  geom_line(col = "black")
```

c) One classical extension in the model is too add in an Allee effect, a minimum threshold density below which experiences a freefall into extinction. This is done by adding in an extra term to the logistic equation above. Can you modify the function above to incorporate the Allee effect. How many equilibria exist in the model now?

```{r Task2c A simple population model- add in the allee effect}
t <- 1:100
K <- 75 # Carrying capacity, range 50-100
r <- 0.3 # Growth rate, range 0-1
alpha <- 20
starting.A <- 10 # runif(1,0,100) # sample the starting value

A <- rep(NA, length(t))
A[1] <- starting.A

dA.dt <- rep(NA, length(t)) # Rate of population growth for timestep 1

for(i in 1: (length(t)-1)) {
  dA.dt[i] <-  r* A[i] * (1 -(A[i]/K) ) *((A[i] - alpha)/K)
  A[i+1] <- A[i]+ dA.dt[i]
  }

A[is.infinite(A)] <- 0

modelResult.allee<- tibble(Time = t, dA.dt = dA.dt, A = A)

popDensityPlot.allee <- modelResult.allee %>%
ggplot(aes(x = Time, y = A)) +
  geom_point() +
  geom_line(col = "black") +
  ylim(0,100)

 growthRate.plot.allee <- modelResult.allee %>%
   ggplot(aes(x = A, y = dA.dt)) +
  geom_point() +
  geom_line(col = "black")

cowplot::plot_grid(popDensityPlot.allee, growthRate.plot.allee, ncol = 1, labels = "auto", align = "hv", axis = "lr")
```

# What are critical transitions?
- Semi-arid systems provide a classical example
- For a given environmental condition, the system can sometimes can be in two different sates
- *Hysteresis* (state at time *t* is dependent on state at *t-1*)
- Critical transitions can result
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img7_path <- "images/semi_arid_feedback.png"
knitr::include_graphics(img7_path)
```

# Hysteresis and critical transitions
- a/b : Only one equilibrium state for a given transition (smooth/ threshold response: e.g. toxicity)
- c Three equilibria (two stable, one unstable), results in two basins of attraction, and a repeller
- Alternative state states, and critical transitions can result (F1/F2 known as *critical bifurcation* points)

**This type of behaviour is often proposed in palaeoecological studies and is something that we would like to detect**
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
knitr::include_graphics(img7_path)
```

# Exercise 3: Alternative equilibria in a consumer-producer system
**Aims**
- To use the same logistic-growth plant population model, but build in an interaction term describing rates of herbivore consumption
- Display properties such as stability, instability, alternative attactors and hysteresis in a theoretical model system, 

## Task one: understading the herbivore consumption model
```{r Task3 Alternative equilibria in a consumer-producer system}
#First, we need to investigate the properties of a model to describe how the rate of consumption changes at different consumption values. This model describes rate of consumption at given levels of food density (i.e. A in the logistic model above). As more food is available, animals will eat more, but the rate of consumption will reach an asymptote when the maximum consumption is reached. At very low levels there will be no consumption because the food supply is difficult to access (e.g because the food is hidden in crevices etc). This is formulated with the Hill equation:

#cons = g * A^p / (A^p + H^p)

#The power p gives the sigmoidal relationship, and g is the maximum food density. H is a half saturation coefficient. Change the parameters of this model in the worksheet to get a feel of how consumption is related to food availability.

g <- 75 # maximum consumption rate, g
H <-  20  # Half saturation rate, H, determines centre point of sigmoidal curve
p <- 2 # power term, gives sigmoidal relationship and determines steepness of the curve
A <- 1:100 # population density of the food supply
cons <- g *A^p / (A^p + H^p)

consumerModel <- tibble (A = A, cons = cons)

conPlot <- consumerModel %>%
   ggplot(aes(x = A, y = cons)) +
  geom_point() +
  geom_line(col = "black") +
  ylim(0,100)

conPlot
```

## Step 2- combining consumption and logistic growth
```{r Task3b Alternative equilibria in a consumer-producer system}
# Now, the logistic growth function of the plants and the consumer model function can be combined. Note that the total consumption is dependent on the number of individuals, so the Hill equation above is multiplied by c (the number of consumers).

# dA/dt = r*A *(1- A/K) - g*c * A^p/ A^p +H^p

# Plant growth parameters
K = 100 # Carrying capacity, range 50-100
r = 0.2 # Growth rate, range 0-1

# Consumer parameters
g = 4 # maximum consumption rate
H = 5 # Sat. rate determines centre point of sigmoidal curve
p = 5  # power term, gives sigmoidal relationship and determines steepness
c =  1   # number of consumer

# First we want to investigate how growth rate of the plants and consumption rate of those plants, changes according to the population of denisty of the plants

A.theoretical <- 1:100 # range of population densities of the plants

dConsumption.dt <- c*g *A.theoretical^p / (A.theoretical^p + H^p)
dA.dt  <-  r*A.theoretical *(1- (A.theoretical/K))

# Now you can plot rates on the same diagram
modelRates <- tibble(Population.density = A.theoretical,
                     consumptionRate = dConsumption.dt,
                     plantGrowthRate = dA.dt)

modelRatesPlot <- modelRates %>%
  pivot_longer(c("consumptionRate", "plantGrowthRate"), names_to = "Model", values_to = "Rate" ) %>%
  ggplot(aes(x = Population.density, y = Rate)) +
  geom_point(aes (col = Model)) +
  geom_line(aes(col = Model))
modelRatesPlot
```

- Now solve the model at different starting population densities using the same parameters. What do you notice happens to the population density of the plants value at different starting values?
- How do these values relate to the `modelRatesPlot` object above?

```{r exploring producer consumer model at different consumption rates}
# Initialise the model and create the output data
t <- 1:100

starting.A <- 10 # change this from 10 to 100
A <- rep(NA, length(t))
A[1] <- starting.A
dA.dt <- rep(NA, length(t)) # Rate of population growth for timestep 1

for(i in 1: (length(t)-1)) {
  dA.dt[i] <-  r* A[i] * (1 -(A[i]/K) )  - (c*g *A[i]^p / (A[i]^p + H^p))
  A[i+1] <- A[i]+ dA.dt[i]
  }

modelResult <- tibble(Time = t, Population.Density = A )

EquilibriumValue <-  round(A[length(A)],1)
popDensityPlot <- modelResult %>%
ggplot(aes(x = Time, y = A)) +
  geom_point() +
  geom_line(col = "black") +
  ylim(0,100) +
 annotate(geom= "text", x=80, y=45, label=paste0("Equil = ", as.character(EquilibriumValue)), color="red")

popDensityPlot
```

Some `R` coding: we could actually turn this into a function so that we can solve this equation automatically and other function to make a plot
```{r consumer-producer full solution}
calculate.Equilibrium <- function(starting.A = 10, parameter.tibble= input.parameters ){

# Set up paramters in the model
g <-   parameter.tibble$g
H <- parameter.tibble$H
p <- parameter.tibble$p
c <- parameter.tibble$c
K <-  parameter.tibble$K
r <- parameter.tibble$r

# Get theoretical growth/ consumption rates for different values of A
A.theoretical <- 1:100 # range of population densities of the plants

dConsumption.dt <- c*g *A.theoretical^p / (A.theoretical^p + H^p)
dA.dt  <-  r*A.theoretical *(1- (A.theoretical/K))

# Now you can plot rates on the same diagram
modelRates <- tibble(Population.density = A.theoretical,
                     consumptionRate = dConsumption.dt,
                     plantGrowthRate = dA.dt)

# Run the model across different timesteps
t <- 1:100
A <- rep(NA, length(t))
A[1] <- starting.A
dA.dt <- rep(NA, length(t)) # Rate of population growth for timestep 1

for(i in 1: (length(t)-1)) {
  dA.dt[i] <-  r* A[i] * (1 -(A[i]/K) )  - (c*g *A[i]^p / (A[i]^p + H^p))
  A[i+1] <- A[i]+ dA.dt[i]
  }

modelResult <- tibble(Time = t, Population.Density = A )
EquilibriumValue <-  A[length(A)]

result <- tibble(EquilibriumValue, modelRates= list(modelRates), modelResult = list(modelResult) )
return(result)

}

# Function to plot the results of the model

plot.consumerModelResult <- function(consumerModelResult) {

  modelRatesPlot <- consumerModelResult$modelRates[[1]] %>%
  pivot_longer(c("consumptionRate", "plantGrowthRate"), names_to = "Model", values_to = "Rate" ) %>%
  ggplot(aes(x = Population.density, y = Rate)) +
  geom_point(aes (col = Model)) +
  geom_line(aes(col = Model))

  EquilibriumValue <-  round(consumerModelResult$EquilibriumValue,1)

popDensityPlot <- consumerModelResult$modelResult[[1]] %>%
ggplot(aes(x = Time, y = Population.Density)) +
  geom_point() +
  geom_line(col = "black") +
  ylim(0,100) +
  annotate(geom= "text", x=80, y=45, label=paste0("Equil = ", as.character(EquilibriumValue)), color="red")

ggpubr::ggarrange(modelRatesPlot, popDensityPlot,
                    ncol = 1, nrow =2)

}
```

Then run the functions to solve the equation for a given starting value and use this function to investigate the effects at different maximum consumption rates
What do you notice about the equilibrium values at different consumption rates?
```{r producer consumer solve}
# Create input parameter tibble
input.parameters <- tibble( # Producer parameters
                            K = 100,  # Carrying capacity, range 50-100
                            r = 0.2, # Growth rate, range 0-1

                          # Consumer parameters
                          g = 4, # maximum consumption rate
                          H = 5, # Sat. rate determines centre point of sigmoidal curve
                          p = 5,  # power term, gives sigmoidal relationship and determines steepness
                          c =  1   # number of consumer
)

result <- calculate.Equilibrium(starting.A = 20, input.parameters)
plot.consumerModelResult(result)
```

# Now we can do this for many different consumption levels
```{r solve full model at different consumption rates}

# Make a function for to estimate equilibria at different starting values
estimate_equilibria <- function(n.consumers = 1) {
  startingValues <- 10:90
  input.parameters$c <- n.consumers
  equilResult <- rep(NA, length(startingValues))

  for(i in 1:length(startingValues)){
    equilResult[i] <-  calculate.Equilibrium(startingValues[i], input.parameters )$EquilibriumValue
  }
  resultTib <- tibble(startingValues = startingValues, EquilibriumResult = equilResult, c = input.parameters$c)
  return(resultTib)
  }

# Run it at different consumption levels
nConsum <- seq(0.1, 2, 0.1)
statesPlot <- do.call(rbind.data.frame, map(nConsum, estimate_equilibria ))
```

Finally we can make a 3D plot of the result
```{r 3D plot}
statesPlot.matrix <- statesPlot %>%
  spread(key = "startingValues", value = "EquilibriumResult") %>%
  as.matrix()
rownames(statesPlot.matrix ) <- statesPlot.matrix[,"c"]
equilibrium <- statesPlot.matrix[,-1]

plotly::plot_ly(x= as.numeric(colnames(equilibrium)),
        y=as.numeric(rownames(equilibrium)),
        z =~ equilibrium

        ) %>%
plotly::add_surface()
```

# Summary from this model exercise
- Used a simple theoretical model to understand alternative stable states in a model system
- Simple feedback relationships can give complex dynamics, both in time series and in terms of relationships with forcing variables
- What we want to do in this workshop is explore these ideas in palaeoecology, but we must always bear in mind there are different types of response relationships that we are interested in (Scheffer and Carpenter 2003)
- In palaeo-data it is unlikely we understand all the forcing variables?
```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img8_path <- "images/scheffer_carpenter_2003.png"
knitr::include_graphics(img8_path)
```

# How can we detect these types of behaviour in empirical data?

# How can we detect these types of behaviour in empirical (palaeo) data?
- Shifts in time-series (Day 2,Day 3)
- Bimodality (Day 4)
- Different response relationships at different values (Day 3?)
- Early warning indicators? (Day 4)

```{r out.width = "30%", fig.align = 'center', echo = FALSE}
img9_path <- "images/scheffer_carpenter_2.png"
knitr::include_graphics(img9_path)
```
